%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Master Thesis 
% Polina Polunina
% November 2022 
%
% License:
% CC-BY-SA 4.0 -- Creative Commons Attribution-ShareAlike 4.0 International
% https://creativecommons.org/licenses/by-sa/4.0/legalcode
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgments}

I am grateful so much to Prof. Dr. Rolf Backofen for giving me the great opportunity to work on this master thesis in his Bioinformatics Group at the University of Freiburg. I thank him for his lectures that gave me a basis to start working on my master thesis. I am also thankful for giving me the opportunity to have a hiwi job in the Bioinformatics group.

I would like to thank Prof. Dr. Wolfgang R. Hess for agreeing to be my second examiner.

Words cannot express my gratitude to Bérénice Batut, my supervisor. She was absolutely dedicated to my work on this thesis, and her contribution is invaluable. Her continuous guidance, advice, code review, proofreading, as well as regular feedback led to the completion of this thesis. In addition to academic support, she also provided me with moral support and motivation, which was crucial in some points of this work. With her supervision, I felt extremely comfortable, believed in myself, and never gave up. I am glad to be thankful to Bérénice for giving me the opportunity to have a hiwi job in the Freiburg Galaxy Team that supported me financially.

I am very thankful to my supervisor Wolfgang Maier. He was always willing to answer my academic questions. His knowledge regarding the topic of my master thesis is so immense that I sometimes got lost, but his explanations were always very understandable. My sincere gratitude goes out to him for his contribution not only with knowledge but also with guidance and ideas throughout the research. It was a pleasure to work with him.

I want to thank everyone from the Freiburg Galaxy Team with their great head Björn Grüning. I was treated with such kindness and helpfulness by them while writing my thesis. In addition to our lunches together, I appreciate our time together during the workdays in the lab. Special thanks are to Cristo and Engy, who regularly checked on my condition, and Mira and Simon, who ensured me always being invited to team activities. 

I am thankful to Teresa Müller for allowing me to have my first hiwi contract and work on an amazing interactive game development project for Street Science Community that involved me in a bio-related topic. This thesis would not have happened without this involvement.

Thanks to Siyu and Yedil, my Street Science Community mates, that never were greedy to cheer me up, which influenced my master project positively.

Many thanks to my study colleagues and friends from the Computer Science program (and not only) at the University of Freiburg for being friendly and supportive. Special thanks are to Karim, Tidiane, Robin, and Roberto for our study sessions in the library, recommendations, and taking care that I do not lose concentration on my thesis and do not burn out.

I want to thank Oleg Aleksandrov, my supervisor at Ural Federal University, and Julia Budchenko, head of the RT Labs affiliate in Yekaterinburg at the time when I worked there. Without their belief in me and their reference, I would not have come to the University of Freiburg.

It is especially important to me to thank my friend Taya for being proud of me and supporting me financially during the hard times I met while writing my master thesis.

Last but not least, I must mention my parents, my sister Olga, and dogs Teddy and Filya. Although they were far away, they supported me financially and morally.